Bow shocks, bow waves, and dust waves

### 2017-12-19 papers/Papers1234/app-dust-wave.tex
Submitted first paper! To MNRAS and arxiv
Finish equations for different regimes in well-coupled case

\newcommand\Qp{\ensuremath{Q_{\text{p}}}}
\newcommand{\grain}{\ensuremath{_{\text{d}}}}
\newcommand{\xsec}{\ensuremath{\sigma\grain}}
\newcommand\frad{\ensuremath{f_{\text{rad}}}}
\newcommand\thm{\ensuremath{\theta_{\text{m}}}}
A dust grain of geometrical cross-section \(\xsec\) situated a
distance \(R\) from a point source of radiation with luminosity \(L\)
will experience a repulsive, radially directed radiative force
\citep[e.g.,][]{Spitzer:1978a}
\begin{equation}
  \label{eq:dust-rad-force}
  \frad = \frac{\xsec \Qp L} {4 \pi R^2 c} e^{-\tau}
\end{equation}
where \Qp{} is the frequency-averaged\footnote{%
  Frequency averages of any quantity \(x\) should be understood as
  weighted by the attenuated source spectrum:
  \(\langle x \rangle_\nu = (L \, e^{-\tau})^{-1} \int_0^\infty x(\nu)\, L_\nu \, e^{-\tau_\nu} \, d\nu
  \).  } %
radiation pressure efficiency\footnote{%
  For absorption efficiency \(Q_{\text{abs}}\), scattering efficiency
  \(Q_{\text{scat}}\), and asymmetry parameter (mean scattering
  cosine) \(g\), we have
  \(\Qp = Q_{\text{abs}} + (1 - g) Q_{\text{scat}}\)
  \citep[e.g., \S~4.5 of][]{Bohren:1983a}.} %
of the grain, \(c\) is the speed of light, and \(\tau\) is the
frequency-averaged optical depth between the source and the grain.
For simplicity, we will consider only the optically thin case,
\(\tau \to 0\).

@@

If \(\frad\) is the only force experienced by the grain, then it will
move on a \textit{ballistic} trajectory, determined by its initial
speed at large distance, \(v_\infty\), and its impact parameter, \(b\).
For \(b = 0\), the grain radially approaches the source with initial
radial velocity \(-v_\infty\), which is decelerated to zero at the distance
of closest approach, \(R_0\), given by energy conservation:
\begin{equation}
  \label{eq:dust-r0}
  R_0 = \frac{\xsec \Qp L} {2 \pi c m\grain v_\infty^2} \ ,
\end{equation}
where \(m\grain\) is the grain mass.  The grain then turns round and
recedes from the source along the same radius, reaching a velocity of
\(+v_\infty\) at large distance.  For \(b > 0\), the trajectory,
\(R\grain(\theta; b)\), is found\footnote{%
  The problem is formally identical to that of Rutherford scattering,
  or (modulo a change of sign) planetary orbits.  The method of
  solution (via introduction of a centrifugal potential term and
  reduction to a 1-dimensional problem) can be found in any classical
  mechanics text \citep[e.g.,][\S~14]{Landau:1976a}.} %
to be hyperbolic, characterized by an eccentricity,
\(\varepsilon = \bigl( 1 + 4 b^2 / R_0^2\bigr)^{1/2}\), and polar angle of
closest approach, \(\thm = \cos^{-1} \varepsilon^{-1}\).  The trajectory is
symmetrical about \(\thm\) and can be written as
\begin{equation}
  \label{eq:dust-r-theta}
  \frac{R\grain(\theta; b)} {R_0} = 
  \frac{ \tfrac12 \bigl( \varepsilon^2 - 1 \bigr)} {\varepsilon \cos(\theta - \thm) - 1} \ , 
\end{equation}
with a total deflection angle of \(2 \thm\), which is equal to
\(90^\circ\) when \(b = 0.5 R_0\).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/test_xyprime_dragoid}
  \caption{Apparent bow shapes in the plane of the sky for
    parallel-stream dragoids as a function of inclination angle.  Drag
    coefficient, \(\alpha\drag\) increases from top-left to bottom-right.
    Inclination \(\abs{i}\) is shown in \(15^\circ\) increments, indicated
    by line color and thickness (see key).}
  \label{fig:dragoid-xy-prime}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/test_xyprime_div_dragoid}
  \caption{As Fig.~\ref{fig:dragoid-xy-prime} but for divergent stream
    dragoids.  Drag coefficient increases from top to bottom, while
    degree of divergence increases from left to right. }
  \label{fig:dragoid-div-xy-prime}
\end{figure}

>>>>

A dust grain of geometrical cross-section \(\xsec\) situated a
distance \(R\) from a point source of radiation with luminosity \(L\)
will experience a repulsive, radially directed radiative force
\citep[e.g.,][]{Spitzer:1978a}
\begin{equation}
  \label{eq:dust-rad-force}
  \frad = \frac{\xsec \Qp L} {4 \pi R^2 c} e^{-\tau}
\end{equation}
where \Qp{} is the frequency-averaged\footnote{%
  Frequency averages of any quantity \(x\) should be understood as
  weighted by the attenuated source spectrum:
  \(\langle x \rangle_\nu = (L \, e^{-\tau})^{-1} \int_0^\infty x(\nu)\, L_\nu \, e^{-\tau_\nu} \, d\nu
  \).  } %
radiation pressure efficiency\footnote{%
  For absorption efficiency \(Q_{\text{abs}}\), scattering efficiency
  \(Q_{\text{scat}}\), and asymmetry parameter (mean scattering
  cosine) \(g\), we have
  \(\Qp = Q_{\text{abs}} + (1 - g) Q_{\text{scat}}\)
  \citep[e.g., \S~4.5 of][]{Bohren:1983a}.} %
of the grain, \(c\) is the speed of light, and \(\tau\) is the
frequency-averaged optical depth between the source and the grain.
For simplicity, we will consider only the optically thin case,
\(\tau \to 0\).

@@

If \(\frad\) is the only force experienced by the grain, then it will
move on a \textit{ballistic} trajectory, determined by its initial
speed at large distance, \(v_\infty\), and its impact parameter, \(b\).
For \(b = 0\), the grain radially approaches the source with initial
radial velocity \(-v_\infty\), which is decelerated to zero at the distance
of closest approach, \(R_0\), given by energy conservation:
\begin{equation}
  \label{eq:dust-r0}
  R_0 = \frac{\xsec \Qp L} {2 \pi c m\grain v_\infty^2} \ ,
\end{equation}
where \(m\grain\) is the grain mass.  The grain then turns round and
recedes from the source along the same radius, reaching a velocity of
\(+v_\infty\) at large distance.  For \(b > 0\), the trajectory,
\(R\grain(\theta; b)\), is found\footnote{%
  The problem is formally identical to that of Rutherford scattering,
  or (modulo a change of sign) planetary orbits.  The method of
  solution (via introduction of a centrifugal potential term and
  reduction to a 1-dimensional problem) can be found in any classical
  mechanics text \citep[e.g.,][\S~14]{Landau:1976a}.} %
to be hyperbolic, characterized by an eccentricity,
\(\varepsilon = \bigl( 1 + 4 b^2 / R_0^2\bigr)^{1/2}\), and polar angle of
closest approach, \(\thm = \cos^{-1} \varepsilon^{-1}\).  The trajectory is
symmetrical about \(\thm\) and can be written as
\begin{equation}
  \label{eq:dust-r-theta}
  \frac{R\grain(\theta; b)} {R_0} = 
  \frac{ \tfrac12 \bigl( \varepsilon^2 - 1 \bigr)} {\varepsilon \cos(\theta - \thm) - 1} \ , 
\end{equation}
with a total deflection angle of \(2 \thm\), which is equal to
\ang{90} when \(b = 0.5 R_0\).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/test_xyprime_dragoid}
  \caption{Apparent bow shapes in the plane of the sky for
    parallel-stream dragoids as a function of inclination angle.  Drag
    coefficient, \(\alpha\drag\) increases from top-left to bottom-right.
    Inclination \(\abs{i}\) is shown in \ang{15} increments, indicated
    by line color and thickness (see key).}
  \label{fig:dragoid-xy-prime}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/test_xyprime_div_dragoid}
  \caption{As Fig.~\ref{fig:dragoid-xy-prime} but for divergent stream
    dragoids.  Drag coefficient increases from top to bottom, while
    degree of divergence increases from left to right. }
  \label{fig:dragoid-div-xy-prime}
\end{figure}

### 2017-12-19 papers/Papers1234/sec-dust-bow-intro.tex
Finish equations for different regimes in well-coupled case

We will mainly consider the canonical case of a bow around a star of
bolometric luminosity, \(L\), with a radiatively driven
wind,\footnote{In \S~\ref{sec:case-inside-out} we consider the case of
  an external luminous source.} which is immersed in an external
stream of gas and dust with density, \(\rho\), and velocity, \(v\).  The
size and shape of the bow is determined by a generalized balance of
pressure (or, equivalently, momentum) between internal and external
sources.  We assume that the stream is supersonic and super-alfvenic,
so that the external pressure is dominated by the ram pressure:
\(\rho v^2\).

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta L / c \ , 
\end{equation}
where \(\eta\) is the momentum efficiency of the wind, which is typically
\(< 1\) \citep{Lamers:1999b}. If the optical depth is very large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  In the single scattering
limit,\footnote{Why multiple scattering is not important.} and
temporarily neglecting the wind, then pressure balance at the bow
apex, a distance \(R\) along the symmetry axis from the star is given by
\begin{equation}
  \label{eq:2}
  \frac{L}{4 \pi c R^2} = \rho v^2 \ ,
\end{equation}
which gives a fiducial bow shock radius in this limit as
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

>>>>

We will mainly consider the canonical case of a bow around a star of
bolometric luminosity, \(L\), with a radiatively driven
wind,\footnote{In \S~\ref{sec:case-inside-out} we consider the case of
  an external luminous source.} which is immersed in an external
stream of gas and dust with density, \(\rho\), and velocity, \(v\).  The
size and shape of the bow is determined by a generalized balance of
pressure (or, equivalently, momentum) between internal and external
sources.  We assume that the stream is supersonic and super-alfvenic,
so that the external pressure is dominated by the ram pressure:
\(\rho v^2\). We first consider the case where the dust grains and gas
are perfectly coupled by collisions (although this assumption is
relaxed below in \S~\ref{sec:imperf-coupl-betw}).

The internal pressure is the sum of wind ram pressure and the
effective radiation pressure that acts on the bow shell.  The
radiative momentum loss rate of the star is \(L/c\) and the wind
momentum loss rate can be expressed as
\begin{equation}
  \label{eq:wind-efficiency}
  \dot{M} V = \eta L / c \ , 
\end{equation}
where \(\eta\) is the momentum efficiency of the wind, which is typically
\(< 1\) \citep{Lamers:1999b}. If the optical depth is very large, then
all of the stellar radiative momentum, emitted with rate \(L/c\), is
trapped by the bow shell.  In the single scattering
limit,\footnote{Why multiple scattering is not important.} and
temporarily neglecting the wind, then pressure balance at the bow
apex, a distance \(R_0\) along the symmetry axis from the star is given by
\begin{equation}
  \label{eq:rad-press-balance-thick}
  \frac{L}{4 \pi c R_0^2} = \rho v^2 \ ,
\end{equation}
which gives a fiducial bow shock radius in this optically thick limit
as
\begin{equation}
  \label{eq:Rstar}
  R_* = \left(\frac{L}{4\pi c \rho v^2}\right)^{1/2} \ .
\end{equation}

We now consider the opposite, optically thin limit.  Although dust
grains typically constitute only a small fraction
\(Z\grain \sim 0.01\) of the mass of the external stream, they
nevertheless dominate the broad-band opacity at FUV, optical and IR
wavelengths if they are present.\footnote{%
  At EUV wavelengths (\(\lambda < \SI{912}{\angstrom}\)), gas opacity
  dominates if the hydrogen neutral fraction is larger than
  \(\approx 0.001\), see discussion of ionization front trapping
  below.} %
If the total opacity (gas plus dust) per total mass (gas plus dust) is
\(\kappa\) (with units of \si{cm^2.g^{-1}}), then the radiative
acceleration in the optically thin limit is
\begin{equation}
  \label{eq:rad-accel}
  a_{\text{rad}} = \frac{\kappa L}{4 \pi c R^2} \ , 
\end{equation}
so that an incoming stream with initial velocity, \(v_\infty\), can be
brought to rest by radiation alone\footnote{%
  For simplicity, we here ignore the effects of pressure gradients and
  shocks, which are important as the velocity approaches the sound
  speed, \(\sound\). In \S~XX below, we show that the resultant
  corrections to \(R_0\) are of order \(\sound / v_\infty\).} %
at a distance \(R_0\) where
\begin{equation}
  \label{eq:rad-poten}
  \int_{R_0}^\infty a_{\text{rad}} \, dr = \tfrac12 v^2 \ , 
\end{equation}
yielding
\begin{equation}
  \label{eq:rad:R0}
  R_0 = \frac{\kappa L}{2\pi c v_\infty^2} \ .
\end{equation}
On the other hand, we can also employ a similar argument to the
optically thick case above by approximating the bow shell as a
surface, and balancing stellar radiation pressure against the ram
pressure of the incoming stream.  The important difference when the
shell is not optically thick is that only a fraction \(1 - e^{-\tau}\) of
the radiative momentum is absorbed by the bow, so that equation~\eqref{eq:rad-press-balance-thick} is replaced with 
\begin{equation}
  \label{eq:rad-press-balance-tau}
  \frac{L (1 - e^{-\tau})}{4 \pi c R_0^2} = \rho v^2 \ .
\end{equation}
In the optically thin limit, \(1 - e^{-\tau} \approx \tau\), so these two
descriptions can be seen to agree so long as
\(\tau = 2 \kappa \rho R_0\), which we will assume to be generally valid.

Then, defining a fiducial optical depth, \(\tau_* = \rho \kappa R_*\), and adding
in the stellar wind ram pressure\footnote{%
  We implicitly assume that the interaction of the stellar wind with
  the external stream can always be treated in the continuum limit.
  This will be true if either the collisional mean free path or the
  ion Larmor radius is much smaller than \(R_0\), which is almost
  always the case.} %
from equation~\eqref{eq:wind-efficiency}, we find that the general bow
radius can be written in terms of the fiducial radius as
\(R_0 = x R_*\), where \(x\) is the solution of
\begin{equation}
  \label{eq:rad-full-x}
  x^2 - \bigl(1 - e^{-2 \tau_* x} \bigr) - \eta = 0 \ .
\end{equation}
Since this is a transcendental equation, \(x\) must be found
numerically, but we can write explicit expressions for three limiting
cases:
\begin{equation}
  \label{eq:x-cases}
  x \approx
  \begin{cases}
    \text{if \(\tau_* \gg 1\):} & (1 + \eta)^{1/2} \\
    \text{if \(\tau_*^2 \ll 1\):} & \tau_* + \bigl( \tau_*^2 + \eta \bigr)^{1/2} \approx
    \begin{cases}
      \text{if \(\tau_*^2 \gg \eta\):} & 2 \tau_* \\
      \text{if \(\tau_*^2 \ll \eta\):} & \eta^{1/2} 
    \end{cases}
  \end{cases}
\end{equation}
The first case, \(x \approx (1 + \eta)^{1/2}\), corresponds to a
\textit{radiation bow shock}; the second case,
\(x \approx 2 \tau_* \), corresponds to a \textit{radiation bow wave}; and the
third case, \(x \approx \eta^{1/2}\), corresponds to a \textit{wind bow shock}.
The two bow shock cases are similar in that the external stream is
oblivious to the presence of the star until it suddenly hits the bow
shock shell, differing only in whether it is radiation or wind that is
providing the internal pressure.  In the intermediate bow wave case,
on the other hand, the external stream is gradually decelerated by
absorption of photons as it approaches the bow.\footnote{A shock does
  still form in this case, but shocked material constitutes a small
  fraction of the total column density of the shell, see \S~XX.}

We now consider the application to bow shocks around main sequence OB stars.
